---
title: "Class 8"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering
Our first example with **kmeans()** function.

```{r}
# Example plot to see how Rmarkdown works
plot(1:10, typ="l")
```

Back to kmeans...

```{r}
 # Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```
```{r}
k<-kmeans(x,centers=2, nstart=20)
k
```

Use the kmeans() function setting k to 2 and nstart=20
 Inspect/print the results
Q. How many points are in each cluster?
```{r}
k$size
```

Q. What ‘component’ of your result object details
     - cluster size?
      -k$size
     - cluster assignment/membership?
```{r}
k$cluster
```
```{r}
table(k$cluster)
```
     
     - cluster center?
```{r}
k$centers
```
     
Plot x colored by the kmeans cluster assignment and
      add cluster centers as blue points
```{r}
palette(c("blue", "red"))
plot(x, col=k$cluster)
points(k$centers, col="green", pch=20, cex=3)
```
      
Q. Repeat for k=3, which has the lower tot.withinss?

```{r}
k3<- kmeans(x, centers=3, nstart = 20)
k3$withinss
k$withinss
```

repeat for k=3 which has lower tot.withinss?
```{r}
k2<- kmeans(x, centers=2, nstart = 20)
k3<- kmeans(x, centers=3, nstart = 20)
k4<- kmeans(x, centers=4, nstart = 20)
k5<- kmeans(x, centers=5, nstart = 20)
k2$tot.withinss
k3$tot.withinss
k4$tot.withinss
k5$tot.withinss
plot(c(k2$tot.withinss, k3$tot.withinss, k4$tot.withinss, k5$tot.withinss))
```

# Hierarchical Clustering in R

Let's try out the **hclust()** function for hierarchical clustering in R. This function needs a distance matrix as input

```{r}
d<-dist(x)
hc<-hclust(d)
plot(hc)
```

```{r}
plot(hc)
# Draws a line on the dendogram
abline(h=8, col="red")

# Cut the tree to yield cluster membership vector

cutree(hc, h=8)
cutree(hc, k=2)
```

### A more 'real' example of data clustering

```{r}
 # Step 1. Generate some example data for clustering
x <- rbind(
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2),   # c1
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
  matrix(c(rnorm(50, mean = 1, sd = 0.3),           # c3
           rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
#         (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
palette(c("red", "blue", "black"))
plot(x, col=col, pch=20)
```

 Q. Use the dist(), hclust(), plot() and cutree()
      functions to return 2 and 3 clusters
```{r}
dist<-dist(x)
clustering2<-hclust(dist)
plot(clustering2)
```

```{r}
#cutree function to return 2 and 3 clusters
grp.3<-cutree(clustering2, h=2)
grp.2<-cutree(clustering2, h=2.5)
```

Q. How does this compare to your known 'col' groups?

```{r}
table(grp.3, col)
```



#Principal Component Analysis

Lets make up some data with quite a few dimensions
```{r}
 ##Initialize a blank 100 row by 10 column matrix
mydata <- matrix(nrow=100, ncol=10)
##Lets label the rows gene1, gene2 etc. to gene100
rownames(mydata) <- paste("gene", 1:100, sep="")
## Lets label the first 5 columns wt1, wt2, wt3, wt4 and wt5
##  and the last 5 ko1, ko2 etc. to ko5 (for "knock-out")
colnames(mydata) <- c( paste("wt", 1:5, sep=""),
                       paste("ko", 1:5, sep="") )
## Fill in some fake read counts
for(i in 1:nrow(mydata)) {
wt.values <- rpois(5, lambda=sample(x=10:1000, size=1)) 
ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))
  mydata[i,] <- c(wt.values, ko.values)
}
head(mydata)
```

Note the precomp()functions want us to take the transpose of our data
```{r}
head(t(mydata))
```
now lets try to find the structure in this data with **prcomp()**

```{r}
pca<-prcomp(t(mydata), scale=TRUE)
pca
```

Make a PC plot of PC2 vs PC2. For this use the '$x' component of output (i.e pca$x)
```{r}
plot(pca$x[,1], pca$x[,2])

```
Lets see how well PCS are doing (how much variance form the original data are they capturing)
```{r}
#variance captured per PC
pca.var<-pca$sdev^2
pca.var
pca.var.per<-round(pca.var/sum(pca.var)*100,1)
pca.var.per
```

Make a "scree-plot" to see the variance captured in each PC
```{r}
barplot(pca.var.per, main="Scree Plot",
        xlab="Principal Component", ylab="Percent Variation")
```

Lets make our plot a bit more useful...
```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```


#Hands on section Food plot failure
```{r}

```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
